The forecast trap

### 2021-09-15 manuscript/manuscript.Rmd
:pen:

---
title: 'The forecast trap'
- name: Carl Boettiger
  email: cboettig@berkeley.edu
  affiliation: a,1
date: "`r Sys.Date()`"

@@

Each possible parameterization of the model will generate a different forecast, and thus require a it's own SDP solution to determine the corresponding policy.
We will consider a larger suite of these logistic models in a moment, but first let us draw out only two such parameterizations from this larger suite.
This will let us focus our attention on two of the most interesting regions of that are already included within that larger parameter space of all possible values for $r$, $K$, and $\sigma$.
Thus, let us take "Model 1" as being given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, "Model 2" by $r_2 = 0.5$, $K_2 = 10$, $\sigma_2 = 0.075$. 
We can imagine our comparison of these two models as a microcosm of the larger comparison between all possible paremeterizations.

@@

Any adaptive management strategy updates posterior distributions over model uncertainty [@Punt2016; @Ludwig1982].
Unfortunately, in this case, any such adaptive updating leads to worse outcomes than the equivalent non-adaptive strategy, in which model uncertainty is held fixed.
I illustrate the application of a passive adaptive management strategy to this simple example, following classic examples for parameter [@Ludwig1982] or structural [@Smith1981] model uncertainty.
Passive adaptive management for a simple sequential decision problem is straightforward to implement over a discrete set of states and actions using dynamic programming with iterative updates [@Smith1981, example code in Appendix].
To demonstrate that the behavior is not driven by failure to explore sufficiently, (which might be addressed by an active adaptive management), I will assign initial probability that model 2 is true at 1%.
After a single iteration of adaptive learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct [Fig 3A].

@@

The forecast trap can occur whenever two conditions are met: 
(1) our candidate models do not include the true model, and
(2) the optimal policy of the true model is not unique to that model.
The first condition is widely accepted as common-place -- ecological processes will always be more complex than our models [@Getz2017].  
The second condition is more subtle: the fact that both curves for model 1 and model 3 produce exactly the same management policy (Fig 4B) under the SDP algorithm may be surprising.
Fortunately, the models we have chosen to focus on here are simple enough to be analytically tractable -- that is, we can solve for the optimal policy without the brute-force numerical approach of SDP.
That solution was first provided by @Reed1979, who demonstrated the optimal policy would be a so-called "bang-bang" strategy, in which the manager seeks to bring the fish stock to some target biomass as quickly as possible whenever it deviates from that size. 
The precise value of that target biomass depends of course on the details of the model -- but importantly not all the details.
Intuitively one can think of this as maintaining the biomass at the most productive size: the maximum population growth rate (position of the peak of the growth curves in Fig 4A), though this is only precisely true without discounting ($\delta = 1$): the optimal stock size $\hat x$ is the solution to $f(\hat x) = \hat x/\delta$ when stochasticity is sufficiently small [@Reed1979].
Thus, all models in which the peak growth rate occurs at the same stock size will have the same optimal policy.
While the SDP is able to discover this same policy numerically by examining the forecast probabilities, Reed's theorem makes it intuitive why that policy is shared between models 1 and 3.

However, Reed's proof considered only one-dimensional growth models $f(x)$, and so tells us nothing about models with stage structure or interacting species.
It is easy to believe that the simple and thus non-unique optimal policy studied by @Reed1979 is an artifact of the simple growth model: surely the intuition provided will not extend to two or more species? 
Fortunately, recent work in the mathematics literature has finally been able to extend Reed's results.
@Hening2021 proves under comparable assumptions that the optimal strategy

<!---
The explanation comes from noticing that the stock size corresponding to the maximum growth rate under model 1 (the peak of the curve) falls at almost exactly the same stock size as that of the peak growth rate for model 3.
Meanwhile, the peak of model 2 occurs at a substantially lower stock size.
While the optimal control solution appears to depend only on a step-ahead forecast accuracy (recall that the SDP solution method used here takes only step-ahead forecast probabilities as input, [@Marescot2013]), mathematical analysis showed long ago [@Reed1979] that the optimal solution for this problem depends only on keeping the biomass at the value responsible for the maximum growth rate.[^1]
Consequently, many models can share the same optimal solution, regardless of their forecast performance.

This result may appear surprising, and begs the question as to whether the non-uniqueness of the optimal solution is not merely an artifact of the simple models considered here. 
Surely more any realistically complex model would require a comparably more complex management policy such that the likelihood that any of our candidate models shares the same optimal policy would be vanishingly small?
The intuition sounds reasonable

[^1]: This is only approximately accurate. @Reed1979 demonstrates 
-->

This realization is quite general: for most decision problems, simple models will exist under which the optimal decision is the same as it would be for the true model, even when that simple model is wrong in most other ways.

<!-- Intuition for it's not just this example -->
Situations similar to this example, in which a model can lead to the optimal decision while producing a poor forecast, are likely to be widespread.

@@

This example demonstrates that it is possible for a model to give quite good forecasts while leading to marginal outcomes, and conversely, that is possible for a model to give optimal outcomes while making terrible forecasts.
This result may be particularly surprising given that the policy is derived directly from the forecast.

In most realistic management scenarios, we do not have the benefit of a theorem similar to Reed's to identify what features are most critical to approximate well and what are not.

For instance, the theorem no longer holds if stock estimates are uncertain [@pomdp-intro].
Absent the guidance of such results, iterative learning over even relatively rich possible candidate models can move us towards worse outcomes than not learning at all.

That potential for long-term deterioration in management outcomes due to adaptive learning has not been widely acknowledged before.
Certainly the limitations of any modeling approach to improve management outcomes due to the human context of competing interests, imperfect enforcement, etc., is readily appreciated [@Ludwig1993].
The problem here goes deeper, to the limits of models and learning from data even in a toy example free from other real world challenges.
Previous work has acknowledged that iterative learning may not always be feasible and that it may not always be beneficial [@Polasky2011], but has failed to recognize the possibility that it can potentially be detrimental.
For example, active adaptive management is premised on the observation that future observations may be too uninformative to distinguish between alternatives [@Walters1978].
Others have also noted that even big improvements in predictive accuracy could have negligible improvement on the decision outcomes, which motivates quantifying the Value of Information (VOI) [@Katz1987].
But in both cases, the worst outcome is wasted effort.
In the scenario considered here, the VOI is negative: any learning over the proposed model uncertainty leads to lower expected net utility than not learning.

Worse, absent the counter-factual of non-adaptive management to compare to, this deterioration in outcomes is invisible. 
All the information available to our manager reinforces the conviction that this deteriorated ecological state is in fact the natural, sustainably managed equilibrium of our fish population.
By responding in any way to future observations to reduce model uncertainty, our manager is caught in a trap of self-fulfilling prophecy:
believing ever more firmly in models that make more accurate predictions while driving decisions that prevent the fish stock from recovering.
The only way to avoid this trap is to _forego learning_ with each new observation, to avoid iterative updates.
It may seem that the adaptive management approach fails because posterior probabilities are updated according to Bayes rule.
Like iterative forecasting, this favors models which better predict the data, despite the fact that under the adaptive management approach, the overall optimization is conditioned on actual utility and not model fit.
However, this issue is not easily avoided.
For example, so-called greedy optimization techniques that favor actions with higher immediate reward do even worse in this context, since such an algorithm would obviously maximize the immediate harvest and therefore collapse the stock.
Only by comparing the net utility derived from management under model 1 for many iterations to the net utility derived under model 2 after many iterations can we determine that model 1 leads to better outcomes.
This raises several questions.
Do such methods of learning and reducing model uncertainty ever lead us towards worse long-term outcomes in the real world? 
If so, How would we know, and what should we do about it?

<!-- **Is it just a unicorn?** -->
Do models which give nearly optimal performance while at the same time making wildly wrong predictions really exist?
It is tempting to argue that model 1 is a kind of unicorn, a mythical creature existing only in theory. But that is a difficult assertion to prove.
Examples of serious forecasts that widely fail to predict future observations abound wherever forecasts are common, from elections to economics to environmental change [e.g. @Tetlock2015].
If there is no shortage of bad forecasts, then could any of them be useful? 
The premise that a model need not perfectly capture reality to be useful is at the very heart of modeling. 
It is also important to note that in many cases, any models being used in decision-making are not necessarily being subjected to the rigorous evaluation and updating steps proposed by adaptive management [@Walters1978] or iterative forecasting [@Dietze2018].
This makes it more likely that such models could persist in practice until now. 
While iterations that revise these models will no doubt improve decision outcomes in many cases, it is worth bearing in mind from the example here that such improvement is not guaranteed.

<!-- So what? -->
If these unicorns do exist in real world management, then what do we do about them?
I believe this is an open question, but that our first step must be to recognize it as such.
We have seen that the problem cannot be resolved by more data alone, and is not the result of over-fitting.
Introducing more models into our candidate set is only helpful to the extent that such models can capture features important to the decision, which as we have seen can be quite different from the feature which matter most to prediction.
We have also seen how not updating our uncertainty estimates, or doing so less frequently, can reveal a unicorn model before it is discounted by further iterations.
Approaches such as iterative forecasting or adaptive management that can reduce model uncertainty over time remain promising and important techniques, and essential strategies to avoid the very real problem of model over-fitting.
But their very success in selecting models which give the most accurate predictions can also be trap.
How that is best done in general is an open question.

>>>>

---
title: "The forecast trap"
name: Carl Boettiger
email: "cboettig@berkeley.edu"
affiliation: a,1
date: "`r Sys.Date()`"

@@

Before this or any other model can generate a forecast, we must first come up with some parameter estimates.
Because the model includes (log-normal) stochastic growth, no amount of data will make any parameter combination impossible, though certain parameter values are more likely than others.
Remember too that parameter estimates may be derived in other ways than than model fitting, especially when parameters are amenable to biological interpretation. 
We will consider the whole range of possible parameters in a moment, but for simplicity, let us begin by focusing in around two of the most interesting regions of that are already included within that larger parameter space of all possible values for $r$, $K$, and $\sigma$.
Let us take "Model 1" as being given by $r_1 = 2$, $K_1 = 16$, $\sigma_1 = 0.05$, "Model 2" by $r_2 = 0.5$, $K_2 = 10$, $\sigma_2 = 0.075$. 
We can imagine our comparison of these two models as a microcosm of the larger comparison between all possible paremeterizations.

@@

Any adaptive management strategy updates posterior distributions over model uncertainty [@Punt2016; @Ludwig1982].
Unfortunately, in this case, any such adaptive updating leads to worse outcomes than the equivalent non-adaptive strategy, in which model uncertainty is held fixed.
I illustrate the application of a passive adaptive management strategy to this simple example, following classic examples for parameter [@Ludwig1982] or structural [@Smith1981] model uncertainty.
Passive adaptive management for a simple sequential decision problem is straightforward to implement over a discrete set of states and actions using dynamic programming with iterative updates [@Smith1981, example code in Appendix].
To demonstrate that the behavior is not driven by failure to explore sufficiently, (active adaptive management), I will assign initial probability that model 2 is true at 1%.
After a single iteration of adaptive learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct [Fig 3A].

@@

@@

It may seem reasonable to expect that the forecast trap would be rare in real world situations:
the chance that the candidate set of models would include anything coming close to the optimal policy of the (unknown) true model without also providing a good forecast seems like it ought to be vanishingly small.
In our example, we were only able to capture the optimal policy with a Gordon-Schafer model because it turned out the optimal policy boiled down to a very simple rule. 
Surely this does not happen in the more complex models of the real world?
Surprisingly, more complex models offer no such guarantee, while real-world constraints make this situation *more* likely, not less.

How can the very different forecasts from model 1 and model 3 could produce exactly the same optimal management policy (Fig 4B) under the SDP algorithm?
Analytic solutions offer more insight as to when and why very different forecasts can generate the identical policy.
Such a solution was first provided by @Reed1979, who demonstrated the optimal policy in the case considered here would be a so-called "bang-bang" policy.
Intuitively one can think of this as maintaining the biomass at the most productive size: the maximum population growth rate (position of the peak of the growth curves in Fig 4A), though this is only precisely true without discounting ($\delta = 1$): the optimal stock size $\hat x$ is the solution to $f(\hat x) = \hat x/\delta$ when stochasticity is sufficiently small [@Reed1979].
Thus, all models in which the peak growth rate occurs at the same stock size will have the same optimal policy.
These are not merely "bad models getting lucky" -- all such models correctly capture the crucial feature relevant to the decision.
In more complex models, such features are more difficult or impossible to identify analytically; but just because we cannot intuit the optimal policy does not mean it is uniquely complex.

Do more complex models lead to more complex control rules?
The theorems of @Reed1979, while quite general, say nothing about structured models or those with predator-prey or competitive interactions.
Yet recent mathematical breakthroughs such as @Holden2015, @Hennig2019, and @Henning2021 have finally been able to extend Reed's theorem to such cases more generally. 
As with Reed's result, these recent proofs make it clear that the optimal strategy for managing these more complex models is not unique, but will be shared by many much simpler models.
So while the true population dynamics may be given by very complex non-linear functions of the interacting species, there will be simple two species models and even un-coupled, one-species models which would lead to the identical optimal policy.
It may be more likely that our candidate set contains a model which matches the optimal policy than that it contains a model which matches the generative process.

Real world considerations actually make this situation more likely, not less likely.
Managers cannot resort to arbitrarily complex policies, regardless of how complex their models.  
Policy adjustments are costly [@Boettiger2016] and the space of available actions is usually far more limited than the space of available states.
In some ecological decision-making contexts, a manager may only be able to select between a handful of alternative actions.
Such constraints make it much more likely that the optimal policy will be replicated by a much simpler model.
Most well-managed marine fisheries are constrained to constant or simple piecewise-linear harvest control rules [@Punt2010]. 
This ensures that an infinite number of possible models will share the optimal solution with the true model.

<!--
Reed's insights illustrate that this happens not because "bad" models sometimes get lucky, but because they capture the key features of the decision process.
This is the goal of all modeling: 
models are simplifying abstractions which capture the details important to the task at hand.
In Reed's case, that key feature is quiet intuitive -- the peak growth rate of the curve (if we ignore discounting). 
That simple intuition does not extend to the more complex cases examined by @Hening2021 and certainly in even richer models where the optimal policy is not tractable.
Consequently, these results do not provide a short-cut for finding an optimal policy without knowing the true model: we cannot merely look for the harvest that generates the peak growth rate (even in Reed's case that would not be optimal due to the role of discounting).
All the same, they are a reminder of why simple, processed based models can lead us to the optimal policy without providing accurate forecasts.
Meanwhile, filtering out or weighing models based on forecast skill _alone_ can precipitate the forecast trap by decreasing any prior belief in such simple but effective models in favor of those with greater forecast skill.
-->

The forecast trap may become a more acute issue as the simple, process-based models that have historically underpinned ecology and conservation policy are challenged by more accurate forecasts from statistical and machine learning tools.  
Buoyed by the rapid expansion of available data and computational power, complex and increasingly opaque models are becoming more common [].
Evaluating such models based on forecast skill will not only reduce concerns about over-fitting, but will make a compelling illustration of their viability as a tool for informing policy.
In many cases, these more accurate forecasts may very well prove invaluable in delivering better (or less bad) real-world outcomes.
But this paper is a reminder that such outcomes are by no means an inevitable consequence of better forecasts.
These more accurate predictions are still not the true model, and it so it is always possible to improve predictive accuracy while less accurately reflecting the unknown key features that really drive the policy decision.
I hope this simple and intuitive logic of Figure 4 will provide a ready reminder as to why the model that produces the best forecast will not always produce the best decision.

<!--
Other fields have already recognized the potential to achieve optimal decisions without the pursuit of ever more accurate predictive models. 
The leading artificial intelligence algorithms used in robotic automation and game-playing [@starcraftii; @atari] no longer attempt to estimate a predictive model which is then used to select an action.
In the past several years, so-called "Model-Free" approaches have come to dominate Deep Reinforcement Learning.
These algorithms have no predictive capacity at all; instead, they try to learn the most effective policy rules directly.
Instead of estimating a predictive model the cutting edge, algorithms which make no attempt to learn the policy directly.
Decision makers do not need to predict the future, they only need to know what to do about it.
-->

<!--

That potential for long-term deterioration in management outcomes due to adaptive learning has not been widely acknowledged before.
Certainly the limitations of any modeling approach to improve management outcomes due to the human context of competing interests, imperfect enforcement, etc., is readily appreciated [@Ludwig1993].
The problem here goes deeper, to the limits of models and learning from data even in a toy example free from other real world challenges.
Previous work has acknowledged that iterative learning may not always be feasible and that it may not always be beneficial [@Polasky2011], but has failed to recognize the possibility that it can potentially be detrimental.
For example, active adaptive management is premised on the observation that future observations may be too uninformative to distinguish between alternatives [@Walters1978].
Others have also noted that even big improvements in predictive accuracy could have negligible improvement on the decision outcomes, which motivates quantifying the Value of Information (VOI) [@Katz1987].
But in both cases, the worst outcome is wasted effort.
In the scenario considered here, the VOI is negative: any learning over the proposed model uncertainty leads to lower expected net utility than not learning.
-->

<!--
Worse, absent the counter-factual of non-adaptive management to compare to, this deterioration in outcomes is invisible. 
All the information available to our manager reinforces the conviction that this deteriorated ecological state is in fact the natural, sustainably managed equilibrium of our fish population.
By responding in any way to future observations to reduce model uncertainty, our manager is caught in a trap of self-fulfilling prophecy:
believing ever more firmly in models that make more accurate predictions while driving decisions that prevent the fish stock from recovering.
The only way to avoid this trap is to _forego learning_ with each new observation, to avoid iterative updates.
It may seem that the adaptive management approach fails because posterior probabilities are updated according to Bayes rule.
Like iterative forecasting, this favors models which better predict the data, despite the fact that under the adaptive management approach, the overall optimization is conditioned on actual utility and not model fit.
However, this issue is not easily avoided.
For example, so-called greedy optimization techniques that favor actions with higher immediate reward do even worse in this context, since such an algorithm would obviously maximize the immediate harvest and therefore collapse the stock.
Only by comparing the net utility derived from management under model 1 for many iterations to the net utility derived under model 2 after many iterations can we determine that model 1 leads to better outcomes.
This raises several questions.
Do such methods of learning and reducing model uncertainty ever lead us towards worse long-term outcomes in the real world? 
If so, How would we know, and what should we do about it?
-->
